A Fast and Tight Heuristic for A* in Road Networks

### 2021-01-21 paper/ch_potentials.tex
improve references

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Nodes represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental sized networks, it is too slow for many applications.
Thus, speed-up techniques have been developed.
One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
They have been used successfully in many real world applications\todo{citation?}.
A CH exploits the inherent hierarchy of road networks.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network during query time.
A further popular example is Multi-Level-Dijkstra~(MLD)~\todo{pre CRP zitat?} also known as CRP~\cite{dgpw-crprn-13}.
It is also used in practice~\todo{citation, bing blog?}.
MLD also also uses shortcut edges.
Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.

@@

Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{}.
A* is a goal-directed variant of Dijkstra's algorithm.
See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
A* uses a \emph{heuristic} to guide the search towards the goal.
A heuristic is function that maps a node $v$ onto an estimate of the distance from $v$ to the goal.
A*'s running time crucially depends on how tight this estimate is.
Further, evaluating the heuristic must be fast.
% The free flow travel time to the goal is a valid heuristic for most features.
% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this paper, we describe CH-Potentials, a heuristic that achieves both goals.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design algorithms that support a multitude of features.
In addition, we describe query optimizations for handling of low-degree nodes, common in road networks.
These low degree optimizations are applicable to Dijkstra's algorithm and A*.

@@

There is a lot of work that extends hierarchical speed-up techniques to more complex settings.
\todo{cite overview}
% For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
MLD/CRP~\cite{dgpw-crprn-13} was developed to allow exchanging weights in reasonable time.
This enable user preferences and live traffic.
MLD/CRP also supports turn costs.
Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH), which also supports exchanging weights and thus supporting user preferences and live traffic.
CH is extended to support turn costs in~\cite{}, CCH in~\cite{bwzz-cchtc-20}.
Traffic predictions are combined with CH in~\cite{swz-sfert-20,bgsv-mtdtt-13}.
\todo{cite correct CATCHUp Version}
\todo{improve! and discuss hierarchical extensions and what they achieve and their difficulties in more details}
Specialized variants for electric vehicle routing are developed in~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited.
Combining all these hierarchical extensions is an unsolved problem.
%
CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, a lot of past works use A* to accelerate hierarchical approaches even further and loose A*'s flexibility.
\todo{cite traffic aware routing}
%

ALT~\cite{gh-cspas-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
Just as our approach, ALT is A*-based.
% However, ALT's heuristic is not preprocessing-tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
% This results in a good but not a preprocessing-tight heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

The performance of A* depends on the tightness of the heuristic.
CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise we could use an unmodified CH).
We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
% When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
\todo{decrease figure height}
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
After $\alpha = 1.1$ the growth slows down.
This illustrates both the strengths and limits of our approach and goal directed search in general.
CH-Potentials can only achieve competitive running times if the application allows for a sufficiently tight lower bounds at preprocessing time.

@@

Table~\ref{tab:building_blocks} depicts the performance of A* with different heuristics and optimizations.
We compare CH-Potentials to three other heuristics.
First, the Zero heuristic where $h(x)=0$ for all nodes $x$.
This corresponds to using Dijkstra's algorithm.
Second, we compare against our own implementation\footnote{
We use 16 landmarks generate with the avoid strategy.
We store all landmark distances consecutive for each node.
This increases cache efficiency when reading many distances for the same node.
We do not disable some landmarks during the query, but use all landmarks all the time.
Also, we do not employ bidirectional search because we want to evaluate and compare the building block CH-Potentials in its simplest and most flexible form.
} of ALT~\cite{}.
Finally we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, that is it is the most efficient heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target node.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
Comparing against Oracle-A* allows us to measure the overhead of the heuristic evaluation of CH-Potentials.
Also, no other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.

>>>>

\section{Introduction}
\label{sec:intro}
The past decade has seen a plethora of research on route planning in large street networks~\cite{bdgmpsww-rptn-16}.
Routing a user through a road network can be formalized as the shortest path problem in weighted graphs.
Nodes represent intersections.
Roads are modeled using edges.
Edges are weighted by their traversal times.
The problem can be solved with Dijkstra's algorithm~\cite{d-ntpcg-59}.
Unfortunately, on continental sized networks, it is too slow for many applications.
Thus, speed-up techniques have been developed.
One popular example are Contraction Hierarchies~(CH)~\cite{gssv-erlrn-12}.
They have been used successfully in many real world applications\todo{citation?}.
A CH exploits the inherent hierarchy of road networks.
In a preprocessing step, additional shortcut edges are inserted, which allow skipping unimportant parts of the network during query time.
A further popular example is Multi-Level-Dijkstra~(MLD)~\cite{swz-umlgt-02} also known as CRP~\cite{dgpw-crprn-13}.
It is also used in practice~\todo{citation, bing blog?}.
MLD also also uses shortcut edges.
Both approaches achieve speed-ups of at least three orders of magnitude over Dijkstra's algorithm.

@@

Our approach decouples extensions from the hierarchical speed-up technique by utilizing the A* algorithm~\cite{hnr-afbhd-68}.
A* is a goal-directed variant of Dijkstra's algorithm.
See Figure~\ref{img:search-space} for an example of nodes traversed during an A* search.
A* uses a \emph{heuristic} to guide the search towards the goal.
A heuristic is function that maps a node $v$ onto an estimate of the distance from $v$ to the goal.
A*'s running time crucially depends on how tight this estimate is.
Further, evaluating the heuristic must be fast.
% The free flow travel time to the goal is a valid heuristic for most features.
% In many settings, it is the tightest lower bound available during preprocessing, i.e., independent of real-time data and user settings.
% However, to be useful we must be able to evaluate the free flow heuristic \todo{ist free flow heuristic besser als tight heuristic?} quickly and exactly.
In this paper, we describe CH-Potentials, a heuristic that achieves both goals.
Internally, the heuristic uses a CH.
Fortunately, this is an implementation detail from the perspective of the A*.
To support a new feature, we only need to modify the A* algorithm.
The heuristic core containing the CH remains untouched.
Extending A* is vastly easier than extending a CH.
This enables us to design algorithms that support a multitude of features.
In addition, we describe query optimizations for handling of low-degree nodes, common in road networks.
These low degree optimizations are applicable to Dijkstra's algorithm and A*.

@@

There is a lot of work that extends hierarchical speed-up techniques to more complex settings.
\todo{cite overview}
% For example, in~\cite{fns-opca-14,gks-rpfof-10} multi-criteria optimization is studied.
MLD/CRP~\cite{dgpw-crprn-13} was developed to allow exchanging weights in reasonable time.
This enable user preferences and live traffic.
MLD/CRP also supports turn costs.
Integrating traffic predictions into CRP was studied in~\cite{bdpw-dtdrp-16}.
In~\cite{dsw-cch-15}, CH is extended to Customizable CH (CCH), which also supports exchanging weights and thus supporting user preferences and live traffic.
CH is extended to support turn costs in~\cite{gv-errnt-11}, CCH in~\cite{bwzz-cchtc-20}.
Traffic predictions are combined with CH in~\cite{swz-sfert-20,bgsv-mtdtt-13}.
\todo{improve! and discuss hierarchical extensions and what they achieve and their difficulties in more details}
\todo{cite traffic aware routing}
Specialized variants for electric vehicle routing are developed in~\cite{DBLP:journals/algorithmica/BaumDPSWZ20,DBLP:conf/aaai/EisnerFS11}.
While these works show that it is possible to extend hierarchical approaches, they also show that it is non-trivial.
Further, in every setting the flexibility available at query time is fairly limited.
Combining all these hierarchical extensions is an unsolved problem.
%
CH-Potentials is not the first work to combine hierarchical approaches and A*~\cite{bdsssw-chgds-10,gkw-blwr-07,bdgwz-sfpcs-19}.
However, a lot of past works use A* to accelerate hierarchical approaches even further and loose A*'s flexibility.
%

ALT~\cite{gh-cspas-05,gw-cppsp-05} and CPD-Heuristics~\cite{DBLP:conf/ijcai/BonoGHS19} are the two techniques with high conceptual similarity to CH-Potentials.
Just as our approach, ALT is A*-based.
% However, ALT's heuristic is not preprocessing-tight.
%
CPD-Heuristics are a combination of A* and Compressed Path Databases (CPD).
A CPD can quickly compute the first edge of a shortest path between any two nodes.
In~\cite{DBLP:conf/ijcai/BonoGHS19}, SRC~\cite{DBLP:conf/socs/StrasserHB14} is used as CPD.
For every distance estimation, a shortest path to the target is computed, whose length is used as the heuristic value.
Unfortunately, the employed CPD's quadratic preprocessing running time is prohibitive on large street networks.
%
MtsCopa~\cite{DBLP:journals/tciaig/BaierBHH15} uses a similar idea to CPD-Heuristics but is only applied in a hunter-prey application.
%
In~\cite{DBLP:conf/ijcai/0002UJAKK18} the weighted graph is embedded into Euclidean space using FastMap such that distances in space and distances in the graph roughly correspond.
The Euclidean distance is then used as A* heuristic.
% This results in a good but not a preprocessing-tight heuristic.
%
The handling of low degree nodes in our A* search is a variation of the techniques used in TopoCore~\cite{DBLP:conf/gis/DibbeltSW15}, which is inspired by~\cite{DBLP:journals/pvldb/FunkeNS14}.

@@

The performance of A* depends on the tightness of the heuristic.
CH-Potentials computes optimal distance estimates with respect to $w_\ell$.
However, for most applications, there will be a gap between $w_q$ and $w_\ell$ (otherwise we could use an unmodified CH).
We evaluate the impact of the difference between $w_q$ and $w_\ell$ on the performance of A*.
The lower bound $w_\ell$ is set to the freeflow travel time.
The query weights $w_q$ are set to $\alpha \cdot w_\ell$, where $\alpha\ge 1$.
% When $\alpha = 1$, the heuristic is tight.
Increasing $\alpha$ degrades the heuristic's quality.
Figure~\ref{fig:scaled_weights} depicts the results.
Clearly, $\alpha$ has significant influence on the running time.
Average running times range from below a millisecond to a few hundred milliseconds depending on $\alpha$.
Up to around $\alpha = 1.1$ the running time grows quickly.
After $\alpha = 1.1$ the growth slows down.
This illustrates both the strengths and limits of our approach and goal directed search in general.
CH-Potentials can only achieve competitive running times if the application allows for a sufficiently tight lower bounds at preprocessing time.

@@

Table~\ref{tab:building_blocks} depicts the performance of A* with different heuristics and optimizations.
We compare CH-Potentials to three other heuristics.
First, the Zero heuristic where $h(x)=0$ for all nodes $x$.
This corresponds to using Dijkstra's algorithm.
Second, we compare against our own implementation\footnote{
We use 16 landmarks generate with the avoid~\cite{gw-cppsp-05} strategy.
We store all landmark distances consecutive for each node.
This increases cache efficiency when reading many distances for the same node.
We do not disable some landmarks during the query, but use all landmarks all the time.
Also, we do not employ bidirectional search because we want to evaluate and compare the building block CH-Potentials in its simplest and most flexible form.
} of ALT~\cite{gw-cppsp-05}.
Finally we compare against a hypothetical \emph{Oracle-A*} heuristic.
This heuristic has instant access to a shortest distance array with respect to $w_\ell$, that is it is the most efficient heuristic possible in our model.
We fill this array before each query using a reverse Dijkstra search from the target node.
Thus, the reported running times of Oracle-A* do \emph{not} account for any heuristic evaluation.
Comparing against Oracle-A* allows us to measure the overhead of the heuristic evaluation of CH-Potentials.
Also, no other heuristic, which only has access to the preprocessing weights, can be faster than Oracle-A*.
